GPU Acceleration of the Material Point Method

### 2019-02-13 pres/pres.tex
pres. finish

\usepackage{textpos}
\usepackage{csquotes}
\usepackage{listings}
\usepackage{xurl}
\urlstyle{sf}
\usepackage{hyperref}
\usepackage{amsmath}
\usepackage{bm}
\usepackage{textcomp}
\usepackage{caption}
\usepackage{color}
\usepackage{subfigure}
\usepackage{wrapfig}
\usepackage[english]{babel}
\usepackage[backend=biber,style=alphabetic]{biblatex}
\usepackage{mathtools}
\usepackage{subdepth}
\usepackage{graphicx}
\usepackage{blindtext}
\usepackage{svg}
\usepackage{xcolor,colortbl}
\usepackage{media9}

@@

\begin{frame}
Sorting can dramatically increase workload performance of subsequent steps for neighboring queries:
\begin{enumerate}
  \item Deep sorted accesses are now \textbf{coalesced}. \label{it:sort_1}
  \item \textbf{Data reuse} due to L2-Cache and/or shared memory. \label{it:sort_2}
\end{enumerate}
\textbf{Double buffer particles} to use last sorted state as input for new sorting to profit from item 1 and 2!
\begin{table}
  \footnotesize
  \begin{tabular}{ | l | r | r | r | r| r |r|}    \hline
    Ordering            &  $\Delta t_c (\mu s)$ & Speedup &VRAM              & SM      & L2     & L2-Hit \\\hline
    Random              &  1,516                & -       &25.0\%            & 3.4\%   & 9.1\%  & 10.8\% \\\hline
    Deep sorted           &    218                & 6.95x   &\textbf{75.3\%}   & 24.4\%  & \textbf{35.0\%} & \textbf{37.8\%} \\
    \hline
 \end{tabular}
\caption{\footnotesize Order dependency of binning of $1024\times1024$ randomly positioned particles in a $128\times128\times128$ grid.}\label{table:bin}
\end{table}
\end{frame}
\subsection{The MPM Specific Transfers}
\begin{frame}
\frametitle{The MPM Specific Transfers}
All MPM operations belong to one of those parallelization schemes:
\begin{itemize}
 \item 1 thread : 1 particle: $\square_p = \square_p \circ \square_p \circ ... \circ \square_p$.
 \item 1 thread : 1 node: $\square_p = \square_p \circ \square_p \circ ... \circ \square_p$.
 \item P2G-transfer: $\square_i = \sum_p \square_p \circ \square_{ip}.$
 \item G2P-transfer: $\square_p = \sum_i \square_i \circ \square_{ip}.$
\end{itemize}
\vfill
MPM-Transfers are executed \textbf{multiple times per physical frame} with varying numbers of variables and mathematical operations.\\\vspace{5}
$\Rightarrow$ Preprocessing steps only need to be done \textbf{once per physical frame}. Sorting already introduced as one of these.
\end{frame}
\section{Delving Deeper: Further Opportunities}

\subsection{d}

\begin{frame}
\frametitle{hi}
  \begin{enumerate}
    \item \cite{MPM:COURSE}
  \end{enumerate}
\end{frame}

>>>>

\usepackage{textpos}
\usepackage{csquotes}
\usepackage{listings}
\usepackage{xurl}
\urlstyle{sf}
\usepackage{hyperref}
\usepackage{amsmath}
\usepackage{bm}
\usepackage{textcomp}
\usepackage{caption}
\usepackage{color}
\usepackage{subfigure}
\usepackage{wrapfig}
\usepackage[english]{babel}
\usepackage[backend=biber,style=alphabetic]{biblatex}
\usepackage{mathtools}
\usepackage{subdepth}
\usepackage{graphicx}
\usepackage{blindtext}
\usepackage{svg}
\usepackage{xcolor,colortbl}
\usepackage{media9}
\usepackage{setspace}

@@

\begin{frame}
Sorting can dramatically increase workload performance of subsequent steps for neighboring queries:
\begin{enumerate}
  \item Deep sorted accesses are now \textbf{coalesced}. \label{it:sort_1}
  \item \textbf{Data reuse} due to L2-Cache and/or shared memory. \label{it:sort_2}
\end{enumerate}
\textbf{Double buffer particles} to use last sorted state as input for new sorting to profit from item 1 and 2!
\begin{table}
  \footnotesize
  \begin{tabular}{ | l | r | r | r | r| r |r|}    \hline
    Ordering            &  $\Delta t_c (\mu s)$ & Speedup &VRAM              & SM      & L2     & L2-Hit \\\hline
    Random              &  1,516                & -       &25.0\%            & 3.4\%   & 9.1\%  & 10.8\% \\\hline
    Deep sorted           &    218                & 6.95x   &\textbf{75.3\%}   & 24.4\%  & \textbf{35.0\%} & \textbf{37.8\%} \\
    \hline
 \end{tabular}
\caption{\footnotesize Order dependency of binning of $1024\times1024$ randomly positioned particles in a $128\times128\times128$ grid.}\label{table:bin}
\end{table}
\end{frame}
\subsection{The MPM Specific Transfers}
\begin{frame}
\frametitle{The MPM Specific Transfers}
All MPM operations belong to one of those parallelization schemes:
\begin{itemize}
 \item 1 thread : 1 particle: $\square_p = \square_p \circ \square_p \circ ... \circ \square_p$.
 \item 1 thread : 1 node: $\square_p = \square_p \circ \square_p \circ ... \circ \square_p$.
 \item G2P-transfer: $\square_p = \sum_i \square_i \circ \square_{ip}.$
 \item P2G-transfer: $\square_i = \sum_p \square_p \circ \square_{ip}.$
\end{itemize}
\vfill
MPM-Transfers are executed \textbf{multiple times per physical frame} with varying numbers of variables and mathematical operations.\\\vspace{5}
$\Rightarrow$ Preprocessing steps only need to be done \textbf{once per physical frame}. Sorting already introduced as one of these.
\end{frame}
\begin{frame}
\begin{minipage}{\textwidth}
  G2P-Transfer: $\square_p = \sum_i \square_i \circ \square_{ip}.$\\
  \includesvg[svgpath =../images/,width=\textwidth]{block_transfer_inv_par}
  Pull-Operations read!
\end{minipage}
\end{frame}
\begin{frame}
Similar to filter or stencil operations on the GPU:
\begin{itemize}
  \item 1 node : 1 thread, split grid into blocks corr. to thread groups.
  \item Interpolation function however dependent on particle position.
  \item Needs to be rerun for every particle in the cell.
\end{itemize}
\vfill
Typical setup of transfers:
\begin{enumerate}
  \item Initialize shared memory (pull: with nodes from global memory).
  \item Perform transfers.
  \item Write back to global memory (push: global atomics since writes on halo).
\end{enumerate}
\end{frame}
\begin{frame}
\begin{minipage}{\textwidth}
  P2G-transfer: $\square_i = \sum_p \square_p \circ \square_{ip}.$\\
  \includesvg[svgpath =../images/,width=\textwidth]{block_transfer_par}
  Pull-Operations read, Push-operations write!\\
  $\Rightarrow$ Race conditions.
\end{minipage}
\end{frame}
\begin{frame}
  Typically simulation domain(grid) much bigger than simulation model. Filtering inactive blocks as a preprocess improves performance.
\begin{itemize}
  \item Block is active if any cell counter is active.
  \item Cell counter is active if it has at least one particle.
\end{itemize}
\vfill
\begin{minipage}{0.9\textwidth}
  \footnotesize
  \includesvg[svgpath =../images/,width=\textwidth]{indirect_dispatch}
\end{minipage}
\end{frame}
\begin{frame}
\begin{itemize}
  \item Block and its halo are always target of shared memory operations $\Rightarrow$ P2G-Pull low occupancy.
  \item Batching multiple particles can increase performance due to hiding synchronization, unroll!
  \item Transfers respect shared memory bank conflicts fully.
  \item Warp divergence for varying cell counts.
\end{itemize}

{\footnotesize
\begin{table}
  \begin{tabular}{ | l | r | r | r | r | r |}    \hline
    Method                 &  $\Delta t_c(\mu s)$ &Speedup         &VRAM  & L2   &SM     \\\hline
    global                 &   44,442             &-               &4.6 \%&34.4\%&7.7\%  \\\hline
    global sorted          &   20,484		  &2.21x           &7.0 \%&\textbf{44.0}\%&16.1\% \\\hline
    P2G-sync              &    \textbf{2,595}    &17.47x          &5.9\%&7.6\%  &\textbf{67.0}\%\\\hline
 \end{tabular}
\end{table}
P2G-transfers of one uniformly million particles with 4 particles per cell with random velocities between $v_x,v_y,v_z \in [-1.0;1.0]$ in a $128\times 128\times128$ grid. Block size is (8,4,4).
}

\end{frame}
\section{Delving Deeper: Further Opportunities}
\subsection{A comparison to \cite{Gao:2018}}
\begin{frame}
  \frametitle{A comparison to \cite{Gao:2018}}

Largely same decision making:\\
\begin{table}
  \begin{tabular}{|l|c|c|}
    \hline
           & Me              & \cite{Gao:2018} \\\hline
    Sort   & Count/Histogram & Count/Histogram \\
	   & for each var.   & sel. variables  \\\hline
    Filtering domain& Filter-op.& Sparse Grid structure \\\hline
    Transfers        & Shared mem. only & Warp-shuffle op.\\\hline
  \end{tabular}
\end{table}
Warp shuffle allows for fast parallel segmented reduction of cells of varying counts.\\
$\Rightarrow$ Solves warp divergence and shared memory issues mostly, thread groups now correspond to particles. Faster in these instances but NVIDIA only.

\end{frame}
